---
title: "Assignment 2: Coding Basics"
author: "Gretchen Barbera"
output: pdf_document
geometry: margin= 1.5 cm
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE} 

install.packages('formatR') 

knitr::opts_chunk$set(tidy.opts=list(width.cutoff=50), tidy=TRUE) 
#received this code from TA to enable all the words to show up
#after knitting my document 

```
## OVERVIEW

This exercise accompanies the lessons/labs in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Canvas.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 55, increasing by fives. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1
I_love_math <- seq(1,55,5) #I named my sequence I_love_math and 
#it is showing up
 #in my environment as the sequence of numbers. 
#I can see my values in the environment but not when I run the code 
seq(1,55,5)
#2
 mean(I_love_math) #this calculates the mean of my sequence that I named I_love_math 
median(I_love_math) #this calculates the median of my sequence 
 
#3
mean(I_love_math) > median(I_love_math) #to see if the mean is greater 
#than the median, I used the greater than symbol to determine 
#if R would give me a true or false value. 
#That True or False value will determine if the mean 
#is greater than the median

```

## Basics, Part 2

5.  Create three vectors, each with four components, consisting of (a) student names, (b) test scores, and (c) whether they are on scholarship or not (TRUE or FALSE).

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
a <- c("PJ", "Teddy", "Poppy", "Simon")
student_names <- c("PJ", "Teddy", "Poppy", "Simon") 
#I am creating the vector student_names for the names of
#the four students that I want to talk about. 
#Having the c in front of the list tells me this is a vector
# vector is characters  

b <-c(100,45,78,86)
test_scores <- c(100,45,78,86) #test scores vector- numbers

c <- c(TRUE,TRUE, FALSE, TRUE)
scholarships <- c(TRUE,TRUE, FALSE, TRUE) 
#scholarships vector-logic

#7 

student_achievers <- data.frame(student_names,test_scores,scholarships)
print(student_achievers) 
#I have combined the vectors into my data frame with the data.frame 
#command and renamed it student_achievers

#8
names(student_achievers) <- c("Student Name","test Score", "On Scholarship")
print(student_achievers)
#renamed the columns accordingly 
#Student names are the names of my family pets. 
#PJ is my cat, so he got a 100

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: 
This data frame has three different variables of data, 
numeric, character, and logic while a matrix 
needs to have all the same type of variables 
(all numeric, all characters, all logic- there can’t be a mix). 

10. Create a function with one input. In this function, use `if`...`else` to evaluate the value of the input: if it is greater than 50, print the word "Pass"; otherwise print the word "Fail". 

11. Create a second function that does the exact same thing as the previous one but uses `ifelse()` instead if `if`...`else `. 

12. Run both functions using the value 52.5 as the input

13. Run both functions using the **vector** of student test scores you created as the input. (Only one will work properly...)

```{r basics 3}
#10. Create a function using if...else
check_pass_fail <- function(score)
{if (score > 50) {print("Pass")} else {print("Fail")}  }

print(check_pass_fail(25)) #practice 

#11. Create a function using ifelse()
check_pass_fail_ifelse <- function(score) {
result <- ifelse(score>50, "Pass", "Fail") }

print(check_pass_fail_ifelse(25)) #practice attempt

print(check_pass_fail_ifelse(52.5)) #practice attempt

#12a. Run the first function with the value 52.5
check_pass_fail(52.5)

#12b. Run the second function with the value 52.5

print(check_pass_fail_ifelse(52.5))

#13a. Run the first function with the vector of test scores 

results <- sapply(test_scores,check_pass_fail)

#I used chatGPT to figure out how to apply the check_pass_fail 
#for all the different test score vectors- 
# it originally did not work but then I used chatGPT and class videos to
#understand what the difference between the two was and how to
#get the function to work 
#(ChatGPT, Personal Communication, September 16, 2024)

#13b. Run the second function with the vector of test scores
#here I had to rename my results so I could get the results 
#through the ifelse function

restults <- ifelse(test_scores > 50, "Pass", "Fail") 
results_ifelse <- ifelse(test_scores > 50, "Pass","Fail")
print(results_ifelse) #enter this code to see the answers
```

14. QUESTION: Which option of `if`...`else` vs. `ifelse` worked? Why? (Hint: search the web for "R vectorization")

> Answer:
They both worked but after some googling.
I needed to use the sapply button to get there for the if.. else function. 
The if... else is not vectorized, which means that it 
can't do multiple vectors at once 
(i.e the logic vector, pass, and the data vector, the test score given ).
I needed to use the sapply for the if.. else to my results 
because of the multiple vectors the code had to 
apply to each of the different elements in the function. The if... else
function is also designed for handling scalar data- not the multiple 
The ifelse worked more effiently for me because 
it was able to asses the different vectors at once 
while if... else could not handle the full vector on its own. (ChatGPT, Personal Communication, Septerber 16, 2024)


**NOTE** Before knitting, you'll need to comment out the call to the function in Q13 that does not work. (A document can't knit if the code it contains causes an error!)