---
title: "Assignment 5: Data Visualization"
author: "Gretchen Barbera"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library (tidyverse)
library (lubridate)
library(cowplot)
library(here)
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
here()

NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed <- read.csv(here("./Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), 
                                                                  stringsAsFactors = TRUE)

NEON_NIWO_Litter_mass_trap_Processed <- read.csv(here("./Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), 
                                                 stringsAsFactors = TRUE)


#2 

class(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$sampledate)
#initially reads as factor format


NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$sampledate <- 
  as.Date(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$sampledate, format = "%m/%d/%y")

class(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$sampledate)
#now it is date 

class(NEON_NIWO_Litter_mass_trap_Processed$collectDate)
#factor originally 

NEON_NIWO_Litter_mass_trap_Processed$collectDate <-
  as.Date(NEON_NIWO_Litter_mass_trap_Processed$collectDate)

class(NEON_NIWO_Litter_mass_trap_Processed$collectDate)
#now it is in "Date" format 
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

my_theme <- 
  theme(
    plot.title = element_text(
      size= 16,
      face= "bold",
      color = "darkblue",
      hjust = 5
    ),
    axis.title.x = element_text(size = 12, face = "italic", color = "darkblue"), 
    axis.title.y = element_text(size = 12, face = "italic", color= "darkblue"),
    axis.text = element_text(size= 12, color = "lightgreen"),
    axis.ticks = element_line(color = "lightgreen"),
    panel.grid.major = element_line(color = "lightblue", size = 0.5),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill= "white"),
    panel.background = element_rect(fill="white"),
    legend.key = element_rect(fill = "white"),
    legend.position = "right",
    complete = TRUE
)


theme_set(my_theme)


```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed, aes(
  x= po4,
  y= tp_ug, 
  color = lakename, 
  shape = lakename
)) +
  geom_point(size= .5) +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(title= "Total Phosphorus By Phosphate",
       x= "Phosphate (PO4)", 
       y= "Total Phosphorus (Tp_ug)") +
  xlim(0.2,1)+
  ylim(0,50)+
  scale_color_manual(factors = c("Peter"= "lightblue", "Paul"= "purple")) + 
  my_theme

#It looks terrible? I need to filter out the NA values because there is a lot of them and I think they are ruining my plot? 

NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed <- NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed %>%
  filter(is.finite(po4) & is.finite(tp_ug))



NTL_clean <- NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed %>%
  filter(!is.na(po4) & !is.na(tp_ug))

NTL_clean$lakename <- as.factor(NTL_clean$lakename)


NTL_clean <- NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed %>%
  filter(is.finite(po4) & is.finite(tp_ug))

po4_by_tpug <- ggplot(NTL_clean, aes(
  x= po4,
  y= tp_ug, 
  color = lakename, 
  shape = lakename
)) +
  geom_point(size= .5) +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(title= "Total Phosphorus By Phosphate",
       x= "Phosphate (PO4)", 
       y= "Total Phosphorus (Tp_ug)") +
  xlim(0,400)+
  ylim(-5,150)

  
print(po4_by_tpug)

#I keep getting the same warning messages to look for infinite data ??

summary(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$po4)
summary(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$tp_ug)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 

BP1<- 
  ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed, aes(
  x= month,
  y= temperature_C, 
  fill= lakename
)) +
  geom_boxplot() +
  labs(title = "Temperature by Month", x= "Month")+
  my_theme
print(BP1)


BP2 <- ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed, aes(
  x= month, 
  y= tp_ug, 
  fill= lakename
)) +
  geom_boxplot()+
  labs(title = "tp_ug Per Month")+
  my_theme
print(BP2)

BP3 <- ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed, aes(
  x= month, 
  y= tn_ug, 
  fill= lakename
)) +
  geom_boxplot()+
  labs(title = "tn_ug Per Month")+
  my_theme
print(BP3)


plot_grid(BP1, BP2, BP3, nrow=3, align = 'v', rel_heights = c(1,1,1))


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 



6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6

needle_litter <- NEON_NIWO_Litter_mass_trap_Processed %>%
  filter(functionalGroup == "Needles")

ggplot(needle_litter, aes(x= collectDate, y= dryMass, color = nlcdClass)) +
  geom_point() +
  geom_line() +
  labs(title= "Dry Mass of Needle Litter By Date ans Class")+
  my_theme



#7



needle_litter <- NEON_NIWO_Litter_mass_trap_Processed %>%
  filter(functionalGroup == "Needles")

ggplot(needle_litter, aes(x= collectDate, y= dryMass)) +
  geom_point() +
  geom_line() +
  labs(title= "Dry Mass of Needle Litter By Date ans Class")+
  facet_wrap(vars(nlcdClass), nrow = 3)+ 
  my_theme




```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
I think that they both have their value. Facets are easier to read and understand seperately but seeing the different nlcdClasses together made it easier to visualize the different between them. The facets are cleaner to look at and I appreciate that they are separated so I can see and interpret the different dryMass values. Plot #6 was is good for comparing values right off the bat and visualizing the difference in values. For myself, plot #7 is better for understanding the data. 