---
title: "Assignment 5: Data Visualization"
author: "Gretchen Barbera"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library (tidyverse)
library (lubridate)
library(cowplot)
library(here)
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
here()

NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed<- 
  read.csv(here("./Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), 
                                                                  stringsAsFactors = TRUE)

NEON_NIWO_Litter_mass_trap_Processed <- 
  read.csv(here("./Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), 
                                                 stringsAsFactors = TRUE)


#2 

#class(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$sampledate)
#initially reads as factor format


NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$sampledate <- 
  as.Date(
    NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$sampledate, format = "%m/%d/%y"
    )

class(
  NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$sampledate
  )
#now it is date 

#class(NEON_NIWO_Litter_mass_trap_Processed$collectDate)
#factor originally 

NEON_NIWO_Litter_mass_trap_Processed$collectDate <- as.Date(
    NEON_NIWO_Litter_mass_trap_Processed$collectDate)

class(NEON_NIWO_Litter_mass_trap_Processed$collectDate)
#now it is in "Date" format 
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

my_theme <- 
  theme(
    plot.title = element_text(
      size= 16,
      face= "bold",
      color = "darkblue",
      hjust = 5
    ),
    axis.title.x = element_text(size = 12, face = "italic", color = "darkblue"), 
    axis.title.y = element_text(size = 12, face = "italic", color= "darkblue"),
    axis.text = element_text(size= 12, color = "lightgreen"),
    axis.ticks = element_line(color = "lightgreen"),
    panel.grid.major = element_line(color = "lightblue", size = 0.5),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill= "white"),
    panel.background = element_rect(fill="white"),
    legend.key = element_rect(fill = "white"),
    legend.position = "right",
    complete = TRUE
)


theme_set(my_theme)


```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
# 
# #ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed, aes(
#   x= po4,
#   y= tp_ug, 
#   color = lakename)) +
#   geom_point(size= .5) +
#   geom_smooth(method = "lm", se = FALSE) + 
#   labs(title= "Total Phosphorus By Phosphate",
#        x= "Phosphate (PO4)", 
#        y= "Total Phosphorus (Tp_ug)", 
# color = "lakename")+
#    coord_fixed(ratio = 1) +
#   xlim(-0.2,400)+
#   ylim(-10,160) +
#   my_theme

#When i did this I got a lot of warning messages that there are NAs/infinite values
#It looks terrible? I need to filter out 
#the NA values because there is
#a lot of them and I think they are ruining my plot?

summary(
  NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$po4
  )
summary(
  NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$tp_ug
  )
#I adjusted the limits based on the values I got from the summary 

 

NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed<- 
  NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed %>%
  filter(
    is.finite(po4) & is.finite(tp_ug)
    )



NTL_clean <- 
NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed %>%
  filter(!is.na(po4) & !is.na(tp_ug))
summary(NTL_clean)

summary(NTL_clean$po4)
#now the NAs are gone and my data is easier to read 

# ggplot(NTL_clean, aes(
#   x= po4,
#   y= tp_ug, 
#   color = lakename)) +
#   geom_point(size= .5) +
#   geom_smooth(method = "lm", se = FALSE) + 
#   labs(title= "Total Phosphorus By Phosphate",
#        x= "Phosphate (PO4)", 
#        y= "Total Phosphorus (Tp_ug)", 
#        color= "Lakename") +
#   xlim(-0.2,400)+
#   ylim(-10,160) +
#     scale_color_manual(values = c( "Paul Lake"= "lightgreen", "Peter Lake"= "blue"))+
#   my_theme

#unique(NTL_clean$lakename)

#why is it so skinny? I am going to change the parameters so it makes my data
#look better- but uses they numbers originally because I looked at the summary
#values for the phospohorus and did based on the max and min- I will readjust 
#so we can see the numbers better

 ggplot(NTL_clean, aes(
  x= po4,
  y= tp_ug,
  color = lakename)) +
  geom_point(size= .5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title= "Total Phosphorus By Phosphate",
       x= "Phosphate (PO4)",
       y= "Total Phosphorus (Tp_ug)",
       color= "Lakename") +
  xlim(-0.2,10)+
  ylim(-5,25) +
    scale_color_manual(values = c( "Paul Lake"= "lightgreen", "Peter Lake"= "blue"))+
  my_theme

 
# summary(NTL_clean$tp_ug)- using the summary functions
 #i readjusted my parameters so we could see my data


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 

BP1<- 
  ggplot(NTL_clean, aes(
  x= month,
  y= temperature_C, 
  fill= lakename
)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Peter Lake" = "blue", "Paul Lake"= "lightgreen")) +
  my_theme
print(BP1)
#I hate the pink and blue so I changed the colors to be specific 
#I also intentionally left off the title of the graph so it would look better with the cowplot

BP2 <- ggplot(NTL_clean, aes(
  x= month, 
  y= tp_ug, 
  fill= lakename
)) +
  geom_boxplot()+
  scale_fill_manual(values = c("Peter Lake" = "blue", "Paul Lake"= "lightgreen"))+
  my_theme
print(BP2)

BP3 <- ggplot(NTL_clean, aes(
  x= month, 
  y= tn_ug, 
  fill= lakename
)) +
  geom_boxplot()+
  scale_fill_manual(values = c("Peter Lake" = "blue", "Paul Lake"= "lightgreen")) +
  my_theme
print(BP3)


plot_grid(BP1, BP2, BP3, nrow=3, align = 'v', rel_heights = c(.7,1.1,1.2))

#it's not the cutest looking bot plot...
#i adjusted the sizes of the boxplots so we can see h
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 
The concentrations of tp_ug and tn_up follow similar trends. They are both higher in Peter Lake and are present between the 7th and 9th months respectively. Looking at the temperature, there between the 7th and 8th months are when we see the temperature increase in Peter Lake. However, the temperatures between Peter and Paul Lakes are similar, there is a higher concentration of both tp_ug and tn_up in Peter Lake. 


6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6

needle_litter <- 
  NEON_NIWO_Litter_mass_trap_Processed %>%
  filter(functionalGroup == "Needles")

ggplot(needle_litter, 
       aes(
         x= collectDate, 
         y= dryMass,
         color = nlcdClass)) +
  geom_point() +
  geom_line() +
  labs(title= "Dry Mass of Needle Litter By Date ans Class")+
  my_theme



#7



needle_litter <- 
  NEON_NIWO_Litter_mass_trap_Processed %>%
  filter(functionalGroup == "Needles")

ggplot(needle_litter, aes(x= collectDate, y= dryMass)) +
  geom_point() +
  geom_line() +
  labs(title= "Dry Mass of Needle Litter By Date ans Class")+
  facet_wrap(vars(nlcdClass), nrow = 3)+ 
  my_theme




```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 
> Answer:
They both have their value. Facets are easier to read and understand separately but seeing the different nlcdClasses together made it easier to visualize the other between them. The facets are cleaner to look at and I appreciate that they are separated so I can see and interpret the different dryMass values. Plot #6 was good for comparing values immediately and visualizing the difference in values. For myself, plot #7 is better for understanding the data.