---
title: "Assignment 9: Spatial Analysis in R"
author: "Gretchen Barbera"
output: html_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on spatial analysis.

## Directions
1. Rename this file `<FirstLast>_A09_SpatialAnalysis.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
4. Work through the steps, **creating code and output** that fulfill each instruction.
5. Be sure to **answer the questions** in this assignment document. Space for your answers is provided in this document and is indicated by the ">" character. If you need a second paragraph be sure to start the first line with ">". You should notice that the answer is highlighted in green by RStudio.
5. When you have completed the assignment, **Knit** the text and code into a single **HTML** file.


## DATA WRANGLING

### Set up your session

1.    Import libraries: tidyverse, sf, leaflet, here, and mapview
2.    Execute the `here()` command to display the current project directory

```{r Prepare workspace, error=F, warning=F}
#1. 
library(tidyverse);library(here);
library(sf);library(mapview);library(leaflet);
library(dplyr);library(ggplot2)

#2. 

here()
getwd()


```


### Read (and filter) county features into an sf dataframe and plot

In this exercise, we will be exploring stream gage height data in Nebraska corresponding to floods occurring there in 2019. First, we will import from the US Counties shapefile we've used in lab lessons, filtering it this time for just Nebraska counties. Nebraska's state FIPS code is `31` (as North Carolina's was `37`).

3.    Read the `cb_2018_us_county_20m.shp` shapefile into an sf dataframe, filtering records for Nebraska counties (State FIPS = 31)
4.    Reveal the dataset's coordinate reference system
5.    Plot the records as a map (using `mapview` or `ggplot`)

```{r Read the county data into an sf dataframe}
#3. Read in Counties shapefile into an sf dataframe, filtering for just NE counties

NE_gage_counties <- st_read(here(
  'Data/Spatial/cb_2018_us_county_20m.shp')) %>%
  filter(STATEFP == '31')

head(NE_gage_counties)
here()

#4. Reveal the CRS of the counties features

glimpse(NE_gage_counties)
st_crs(NE_gage_counties)

st_crs(NE_gage_counties)$epsg


#5. Plot the data

ggplot(data = NE_gage_counties) +
  geom_sf(fill = "lightblue", color = "darkblue")+
  labs( title = "Nebraska Counties Map")


mapview(NE_gage_counties)



```

6.    What is the EPSG code of the Counties dataset? Is this a geographic or a projected coordinate reference system? (Om other words, does this CRS use angular or planar coordinate units?) To what datum is this CRS associated? (Tip: look for the EPSG code on <https://epsg.io>)

> ANSWER: 
> The EPSG is 4269. This is a geographic coordinate referense system because it uses degress of lattitude and longitude to describe the gage locations. The CRS uses angular units. According to the site above, the datum is the North American Datum 1983.


### Read in gage locations csv as a dataframe, then display the column names it contains

Next we'll read in some USGS/NWIS gage location data added to the `Data/Raw` folder. These are in the `NWIS_SiteInfo_NE_RAW.csv` file.(See `NWIS_SiteInfo_NE_RAW.README.txt` for more info on this dataset.)

7.    Read the `NWIS_SiteInfo_NE_RAW.csv` file into a standard dataframe, being sure to set the `site_no` field as well as other character columns as a factor. 

8.    Display the structure of this dataset.

```{r Read in USGS Gage Locations}
#7. Read in gage locations csv as a dataframe

gage_locations <- read.csv(here('./Data/Raw/NWIS_SiteInfo_NE_RAW.csv'),
                          stringsAsFactors = TRUE)

head(gage_locations)


#8. Display the structure of the dataframe

str(gage_locations)


```

9.    What columns in the dataset contain the x and y coordinate values, respectively?  
> ANSWER:
> Lattitude and longitude are the x and the y values (we talked about this in class/over the videos). X is longitude- or dec_long_va, and Y is the lattitude- or dec_lat_va. 


### Convert the dataframe to a spatial features ("sf") dataframe

10.   Convert the dataframe to an sf dataframe.
    *   Note: These data use the same coordinate reference system as the counties dataset

11. Display the structure of the resulting sf dataframe

```{r Convert dataframe to sf dataframe}
#10. Convert to an sf object

gage_locations_sf <- gage_locations %>%
  st_as_sf(coords = c("dec_long_va", "dec_lat_va"),
           crs=4269)



#11. Display the structure


str(gage_locations_sf)
st_crs(gage_locations_sf)
#when I do this i want to make sure i get the same information that i got when looking on the website above to make sure that I am using the correct points 
```

12.   What new field(s) appear in the sf dataframe created? What field(s), if any, disappeared?

>ANSWER: 
>The sf data object gained the geometry field which stores the spatial information for the gages. The lat and long columns are not longer in the sf dataset. This is now a spatial dataframe that can be used for spatial analysis and visualizations- I can make maps with the gages now. 


### Plot the gage locations on top of the counties

13. Use `ggplot` to plot the county and gage location datasets.
  *   Be sure the datasets are displayed in different colors
  *   Title your plot "NWIS Gage Locations in Nebraska"
  *   Subtitle your plot with your name

```{r Plot the spatial features}
#13. Plot the gage locations atop the county features


ggplot()+
  geom_sf(data = NE_gage_counties,
          fill= "lightblue", color = "darkblue")+
  geom_sf(data = gage_locations_sf, color = "orange")+
  labs(
    title = "NWIS Gage Locations in Nebraska",
    subtitle = "Gretchen Barbera")

#when I originally ran the plot it said I can't transform the object with the missing crs values so I had to go back up to when I was making the gage_locations_sf to add the CRS values 

```


### Read in the gage height data and join the site location data to it.

Lastly, we want to attach some gage height data to our site locations. I've constructed a csv file listing many of the Nebraska gage sites, by station name and site number along with stream gage heights (in meters) recorded during the recent flood event. This file is titled `NWIS_SiteFlowData_NE_RAW.csv` and is found in the Data/Raw folder.

14. Read the `NWIS_SiteFlowData_NE_RAW.csv` dataset in as a dataframe
    *   Pay attention to which fields should be imported as factors!
15. Show the column names.
16. Join our site information (already imported above) to these gage height data
    *   The `site_no` and `station_nm` can both/either serve as joining attributes
    *   Construct this join so that the result only includes records features where both tables have data (N=136)
17. Show the column names of this resulting spatial dataframe
18. Show the dimensions of the resulting joined dataframe

```{r Read in gage site data}
#14. Read the site flow data into a data frame

site_flow_data <- read.csv(here("Data/Raw/NWIS_SiteFlowData_NE_RAW.csv"))

head(site_flow_data)

str(site_flow_data)


site_flow_data$site_no <- as.factor(site_flow_data$site_no)
#site flow data should be read as factors. I originally has stringsAsFactors = true but then it changed some of the columns that I didn't want to read as factors so I went back to convert


# Check the structure again
str(site_flow_data)


#15. Show the column names

colnames(site_flow_data)
colnames(gage_locations_sf)

#before joining- I have to make sure they have the same site_no column

#16. Join the flow data to our NWIS gage location spatial dataframe


#ran into an error where the site_no were not the same- one was an integer and one was a factor 

site_flow_data$site_no <- as.integer(as.character(site_flow_data$site_no))

gage_data_joined <- gage_locations_sf %>%
  left_join(site_flow_data, by = 'site_no')

#now site_no is an integer for both
#we joined the gage locations and the gage_ht information

#17. Show the column names in the resulting spatial dataframe

colnames(gage_data_joined)

#18. Show the dimensions of this joined dataset

dim(gage_data_joined)

```

### Map the pattern of gage height data

Now we can examine where the flooding appears most acute by visualizing gage heights spatially.
19.   Plot the gage sites on top of counties (using `mapview`, `ggplot`, or `leaflet`) 
    *   Show the magnitude of gage height by color, shape, other visualization technique.

```{r map gage sites}
#19 Map the points, sized by gage height

ggplot() +
  geom_sf(data = NE_gage_counties, fill ="lavender", color = "darkblue") +
  geom_sf(data = gage_data_joined, aes(color = gage_ht, size = gage_ht)) +
  labs (
    title = "Flood Gage Heights in Nebraska"
  )
       
mapview(NE_gage_counties) + 
  mapview(gage_data_joined, 
          zcol = "gage_ht",
          cex = "gage_ht",
          color = "yellow", 
          popup = TRUE, 
          shape = "circle",
          layer.name = "Gage Heights")

colnames(gage_data_joined)
 
#I originally liked the ggplot better but I think the mapview looks best for this 
#adjusted the size of the points based on the gage_ht 
```

---

## SPATIAL ANALYSIS

Up next we will do some spatial analysis with our data. To prepare for this, we should transform our data into a projected coordinate system. We'll choose UTM Zone 14N (EPGS = 32614).


### Transform the counties and gage site datasets to UTM Zone 14N

20. Transform the counties and gage sf datasets to UTM Zone 14N (EPGS = 32614).
21. Using `mapview` or `ggplot`, plot the data so that each can be seen as different colors

```{r spatial analysis}
#20 Transform the counties and gage location datasets to UTM Zone 14

gage_locations_sf_utm <- st_transform(gage_locations_sf, crs = 32614)

NE_gage_counties_utm <- st_transform(NE_gage_counties, crs = 32614)

#i originally did the whole lab using these UTM datasets and then I realized
#I joined them together into one already and it allowed me to keep the gage_ht


gage_data_joined_utm <- st_transform(gage_data_joined, crs = 32614)
  

str(gage_data_joined_utm)
colnames(gage_data_joined_utm)

#21 Plot the data

mapview(gage_data_joined_utm,
        zcol = "gage_ht",             
        cex = "gage_ht",               
        col.regions = c("red", "yellow", "blue")) 


ggplot() +
  geom_sf(data = NE_gage_counties_utm, fill = "lightblue", color = "black", size = 0.1) +  
  geom_sf(data = gage_data_joined_utm, aes(color = gage_ht, size = gage_ht)) +           
  scale_color_gradientn(colors = c("pink", "red"), name = "Gage Height") +     
  scale_size_continuous(range = c(1, 5), name = "Gage Height") +                        
  labs(
    title = "Gage Heights Across Nebraska",
    subtitle = "Symbolized by Gage Height (color and size)"
  ) +
  theme_minimal()

     

#I like the ggplot better for this one 
#make the gages different colors 

```


### Select the gages falling within a given county

Now let's zoom into a particular county and examine the gages located there. 
22. Select Lancaster county from your projected county sf dataframe 
23. Select the gage sites falling within that county to a new spatial dataframe
24. Select the gage sites within 15km of the county to a new spatial dataframe
25. Create a plot showing (each symbolized distinctly):
 * all Nebraska counties,
 * the selected county, 
 * the gage sites in that county,
 * and the gage sites within 15 km of the county


```{r spatial selection}
#22 Select the county

lancaster_county <- NE_gage_counties_utm %>%
  filter(NAME == "Lancaster")

print(lancaster_county)
mapview(lancaster_county)

#23 Spatially select gages within the selected county

gages_in_lancaster <- gage_data_joined_utm %>%
  filter(st_within (geometry, lancaster_county$geometry, sparse = FALSE) [,1])

colnames(gage_locations_sf_utm)

#the st_within filters the gages to keep them in lancaster 
#geometry uses the location of lancaster 
#and then the the lancaster_county$geometry 
#gets the location of the gage points within the county 

print(gages_in_lancaster)

mapview(gages_in_lancaster)


#i mapped the county to see what it looked like- just a regular old rectangle 

#24 Spatially select gages within 15 k of the selected county

lancaster_buffer <- st_buffer(lancaster_county, dist= 15000) 

print(lancaster_buffer)
mapview(lancaster_buffer)

lancaster_gages_within_15k <- gage_locations_sf_utm %>% filter(st_within(geometry, lancaster_buffer, sparse = FALSE))
                                                  mapview(lancaster_gages_within_15k)

print(lancaster_gages_within_15k)



#25 Plot

colnames(lancaster_gages_within_15k)


#mapview(NE_gage_counties_utm, color = "green") + 
  # mapview(gage_data_joined_utm, zcol="gage_ht",
  #         cex= "gage_ht",
  #         col.regions = c("red","blue")) + 
 # mapview(lancaster_buffer, col.regions = "pink")+
 # mapview(lancaster_gages_within_15k)



# ggplot() +
#   geom_sf(data = NE_gage_counties_utm, fill = "gray", color = "black") + 
#   geom_sf(data = lancaster_county, fill = "lightblue", color = "blue") +                     
#   geom_sf(data = lancaster_county, color = "red") +                        geom_sf(data = lancaster_gages_within_15k, color = "purple") +                                 labs(
#     title = "Nebraska Counties, Selected County, and Gage Sites",
#     subtitle = "Gage sites within the selected county and within 15 km of its border")



colnames(lancaster_county)


mapview(NE_gage_counties_utm, color = "black", layer.name = "All Nebraska Counties") +
  mapview(lancaster_county, color = "blue", layer.name = "Lancaster County") +
  mapview(lancaster_county, col.regions = "red", cex = 3,
          layer.name = "Gages in Lancaster County") +
  mapview(lancaster_gages_within_15k, color = "yellow", 
          cex = 3, 
          layer.name = "Gages within 15 km of Lancaster County")+
  mapview(lancaster_buffer, color = "pink")

#this one has a better view of the different pieces

```
