---
title: 'Assignment 8: Time Series Analysis'
author: "Gretchen Barbera"
date: "Fall 2024"
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

```{r}
library(ggplot2)
library(tidyverse)
library(lubridate)
library(trend)
library(zoo)
getwd()

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)


```


2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r, message = FALSE}
#1

# Load required libraries
library(tidyverse)
library(lubridate)
library(zoo)
library(trend)


# If I want to import everything from my specific folder #I have to go to the path 

folder_path <- "~/EDE_Fall2024/Data/Raw/Ozone_TimeSeries"
file_list <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)

GaringerOzone <- file_list %>%
  map_dfr(read_csv)

```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
# 3

GaringerOzone$Date <- as.Date(GaringerOzone$Date, format="%d/%m/%Y")  
 str(GaringerOzone$Date)
 
class(GaringerOzone$Date)




# 4

library(dplyr)

GaringerOzone1 <- GaringerOzone %>%
  select(Date, `Daily Max 8-hour Ozone Concentration`, DAILY_AQI_VALUE)

str(GaringerOzone)

# 5

Days <- as.data.frame(seq(from = as.Date("2010-01-01"), to= as.Date("2019-12-31"), by ="day"))
              

colnames(Days) <- "Date"

str(Days)
head(Days)
colnames(Days)


# 6

GaringerOzone <- left_join(Days,GaringerOzone,by = "Date")

dim(GaringerOzone)




#I had to go back and rename my firts GaringerOzone dataframe and then rename my subset 

```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r}
#7

library(ggplot2)



ggplot(GaringerOzone, aes(x= Date, y= `Daily Max 8-hour Ozone Concentration`)) +
  geom_line(color = "black", size= 0.5)+
  geom_smooth(method = "loess", se= FALSE, color= "pink")





```

>Answer:
Yes, the trend shows a slight decrease in the daily maximum 8-hour ozone concentration over from 2010 to 2014 and then levels out from 2015 to 2020. This trends shows that ozone levels are decreasing during early data collection and then follow a more stablized period. The vertical lines suggest a day to day variability throughout the years which could explained by seasonal variability or weather or other variables not discussed in the plot. 

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8
#install.packages("zoo")
library(zoo)

GaringerOzone$Date <- as.Date(GaringerOzone$Date)

# Perform linear interpolation for the 'Daily Max 8-hour Ozone Concentration' column
GaringerOzone$`Daily Max 8-hour Ozone Concentration` <- na.approx(
  GaringerOzone$`Daily Max 8-hour Ozone Concentration`,
  x = GaringerOzone$Date,
  na.rm = FALSE
)

# Check for any remaining missing values after interpolation
sum(is.na(GaringerOzone$`Daily Max 8-hour Ozone Concentration`)) 

# Display structure and first few rows to verify changes
str(GaringerOzone)
head(GaringerOzone)

# Verify that all dates are intact
any_missing_dates <- any(is.na(GaringerOzone$Date))
print(any_missing_dates)


#the na.approx is used for linear interpolation 
#this just means that it fills in the missing information based on the surrounding data- assuming there are no outliers 
```

> Answer: 
Since 2010, ozone concentrations have declined and then leveled off around 2014 to continue to stay relatively level to when the dataset concludes. There have been individual changes to the ozone levels, but statistcially, the ozone levels have stayed relatiely the same after the initial decline from 2010-2014. 

9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r}
#9

GaringerOzone.monthly <- GaringerOzone %>%
  mutate(year= year(Date),
        month= month(Date) 
         ) %>%
  group_by(year, month) %>%
#year and month columns have been added
  summarize(mean_ozone = mean(`Daily Max 8-hour Ozone Concentration`, na.rm = TRUE), .groups = 'drop') %>%
  mutate(
    Date = make_date(year, month, 1))


#now I have just the first day of the month included with the date


GaringerOzone.daily <- GaringerOzone %>%
  mutate(year= year(Date), 
         month=month(Date),
         day=day(Date)) %>%
  group_by(year, month, day) %>%
  summarize(mean_ozone=mean(`Daily Max 8-hour Ozone Concentration`, na.rm = TRUE),
            .groups = 'drop') %>%
  mutate(
    Date = make_date()
  )




```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r}
#10


GaringerOzone.daily.ts <- ts(GaringerOzone.daily$Date, start = c(year(min(GaringerOzone.daily$Date)), yday(min(GaringerOzone.daily$Date))),
                             frequency = 365)

print(GaringerOzone.daily.ts)


#I have to also get the monthly 

GaringerOzone.monthly$Date <- as.Date(paste0(GaringerOzone.monthly$Date, "-01"))


# Create a monthly time series object
GaringerOzone.monthly.ts <- ts(GaringerOzone.monthly$Date, start = c(year(min(GaringerOzone.monthly$Date)),
                                                            yday(min(GaringerOzone.monthly$Date))),
                    frequency =12)

print(GaringerOzone.monthly.ts)


```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r}
#11

GaringerOzone.daily_decomposed <- stl(GaringerOzone.daily.ts, s.window = "periodic")

plot(GaringerOzone.daily_decomposed)
 





```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12

ozone_trend <- trend::smk.test(GaringerOzone.monthly.ts)

ozone_trend

summary(ozone_trend)

```

> Answer: 
A montonic trend analysis show a gradualshift over time in a consistent direction. A Mann-Kendall test is the most appropriate in this case becauseit is designed to detect trends in seasonality. Because we are looking at the different months, the Kendall-Mann will test the trends each month independently comparing only the monthly data with the specific month (June data is compared to other June data). From the trend that we saw in question #7, the trend is not linear and the Kendall-Mann test does not need the trend to be linear or normally distributed.

(Information collected from class notes and https://real-statistics.com/time-series-analysis/time-series-miscellaneous/mann-kendall-test/#:~:text=The%20Mann%2DKendall%20Test%20is,that%20there%20is%20no%20autocorrelation.) 

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r}
# 13

ggplot(GaringerOzone.monthly, aes(x = month, y = mean_ozone)) +
  geom_line(color = "blue", linewidth = 1) +   
  geom_point(color = "pink", size = 2) +
  labs(
    title = "Mean Monthly Ozone Concentrations Over Time",
    x = "Time (Month)",
    y = "Ozone Concentration"  
  ) +
  mytheme                       



```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: 
The plot shows the ozone concentration fluctuation by month, suggesting that there is relationship between the months and the ozone concentrations levels. There is not a constant or linear trend analysis to this relaitonship.


15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r}
#15


GaringerOzone.monthly.decomp <- decompose(GaringerOzone.monthly.ts)

plot(GaringerOzone.monthly.decomp)


#16


seasonal_component <- GaringerOzone.monthly.decomp$seasonal


non_seasonal_ozone <- GaringerOzone.monthly.ts - seasonal_component


non_seasonal_trend <- trend::smk.test(non_seasonal_ozone)

non_seasonal_trend
summary(non_seasonal_trend)

library(Kendall)

non_seasonal_ozone <- MannKendall(non_seasonal_ozone)

print(non_seasonal_ozone)



```

> Answer: 
After taking the seasonality out of the dataset, the MannKendall results show a <2.22e-16 p-value which means that we can reject the null hypothesis and accept that there is a signficant trend in the data. This aligns with the data we looked at earlier that also had a tau of 1 and a low p-value, indicuting that there is a significant trend between ozone concentrations and months- which can also be visualized in the plot of mean ozone values per month (Q13). 

